{\steve}: A hierarchical Bayesian model for Supernova Cosmology

### 2019-03-11 papers/methods/paper.tex
Adding 'computationally'

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the \added{ possible} events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the denominator represents an integral over all potential data $D$. This is derived in Appendix \ref{app:selection1}. \added{ To evaluate the effect of our selection effects, we need to evaluate both the selection effect terms in the numerator and the integral in the denominator. The numerator represents the probability we caught the supernova and it was selected into the cosmology sample. The integral represents our global selection efficiency at a location in parameter space, rather than the probability of our data being selected into our sample.} As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L}(\theta; {\rm data}) = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) } \label{eq:general1}
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. \added{ At this point, our derivation depends on whether our selection effects are best modelled as a function of observables or as a function of latent parameters. In most cases, selection effects will be best modelled directly as a function of observables, and we would introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L}(\theta; {\rm data}) &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}. \label{eq:app_main}
\end{eqnarray}
The second option, wherein selection effects can be more efficiently modelled with the inclusion of latent parameters (for example, in the case where we do not have direct access to the observables upon which our data selection is determined), we can introduce our latent parameters in addition to an integral over all possible data:
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{\int P(S| L,\theta) P({\rm data}|L) P(L|\theta)\, dL}{\iint P(S|D,L,\theta)P(D|L,\theta) P(L|\theta)\, dD\, dL},\label{eq:latent}
\end{eqnarray}
where $L$ represents our latent parameters that model the true underlying values of our observables, such that our data is conditioned directly on them.
}

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
Here we can now move from the generic labels $D$ and $L$ to something specific to our model. In this case, $D$ would represent our observables, $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{m} \rbrace$, and $L$ our latent parameters $\lbrace m_B, x_1, c, z\rbrace.$ We assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and due to this we will not write out the other parameters in our model for simplicity. If we did, we would simply find them integrated to unity. Note that this assumption does not capture biases engendered by
Poisson noise fluctuations.  
We advocate that future analyses with
higher statistical precision do use a precisely determined $P(S|D)$.
Formally, we assume that $P(S|D, L, \theta) = P(S|z) P(S|m_B)$.}
Writing out the relevant observables $\hat{m}_B$, $\hat{z}$ and latent parameters $m_B$ and $z$, our denominator then becomes
\begin{equation}
d = \int P(S|z) P(S|m_B) P(\hat{z}|z) P(\hat{m}_B|m_B) P(z, m_B | \theta)\, d\hat{z} \, d\hat{m}_B \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
Due to our assumed perfect measurement of redshift, $P(\hat{z}|z) = \delta(\hat{z} - z)$. $P(\hat{m}_B | m_B)$ is a Gaussian due to our Gaussian model of summary statistics, and can be analytically integrated out, collapsing the integral over $\hat{m}_B$. Finally, we can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.

>>>>

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the \added{ possible} events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, 
\end{align}
where the denominator represents an integral over all potential data $D$, \added{and $\theta$ represents top-level parameters. In the case that our selection effects are best charactersied by latent variables \textit{instead of} data, we can add them to our formulation and our likelihood becomes
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{\int P(S|L,\theta) P({\rm data}|L) P(L|\theta)\, dL}{\iint P(S|L,\theta) P(L|\theta) dL},\label{eq:main}
\end{eqnarray}
where $L$ represents our latent parameters.}
This is derived in Appendix \ref{app:selection1}. \added{ To evaluate the effect of our selection effects, we need to evaluate both the selection effect terms in the numerator and the integral in the denominator. The numerator represents the probability we caught the supernova and it was selected into the cosmology sample. The integral represents our global selection efficiency at a location in parameter space, rather than the probability of our data being selected into our sample.} As $\theta$ represents the vector of all \added{top-level model parameters, and $L$ represents a vector of all latent parameters, this is not a trivial integral.} Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L}(\theta; {\rm data}) = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) } \label{eq:general1}
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. \added{ At this point, our derivation depends on whether our selection effects are best modelled as a function of observables or as a function of latent parameters. In most cases, selection effects will be best modelled directly as a function of observables, and we would introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L}(\theta; {\rm data}) &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}. \label{eq:app_main}
\end{eqnarray}
The second option, wherein selection effects can be more computationally efficiently modelled with the inclusion of latent parameters (for example, in the case where we do not have direct access to the observables upon which our data selection is determined), we can introduce our latent parameters in addition to an integral over all possible data:
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{\int P(S|{\rm data}, L,\theta) P({\rm data}|L) P(L|\theta)\, dL}{\iint P(S|D,L,\theta)P(D|L,\theta) P(L|\theta)\, dD\, dL},\label{eq:latent}
\end{eqnarray}
where $L$ represents our latent parameters that model the true underlying values of our observables, such that our data is conditioned directly on them. In this formulation, the selection effects can depend both on data and latent variables.
}

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
In our formulation, we assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. This allows us to isolate our integral $\int P(D|L,\theta)$ in the above equation, integrate it to unity, and come to 
\begin{equation}
d = \int  P(S|L) P(L | \theta)\, dL.
\end{equation}	
Here we can now move from the generic label $L$ to something specific to our model. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and so $L \rightarrow \lbrace m_B, z\rbrace$, or formally $P(S|D, L, \theta) = P(S|z) P(S|m_B)$. We do not write out other latent parameters found in the model as they do not impact selection effects --- we would simply find them integrated to unity. Note that this assumption does not capture biases engendered by
Poisson noise fluctuations.  
We advocate that future analyses with
higher statistical precision do use a precisely determined $P(S|D)$.
Writing out the latent variables, our denominator becomes
\begin{equation}
d = \int P(S|z) P(S|m_B)P(z, m_B | \theta)\, d\hat{z} \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
We can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state}
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.
